\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-02-23 trunk/arch.tex
Rewrote some parts of related.tex with comments

\section{Planning and execution in the \rx architecture}
\label{sec:arch}

% {\em\tiny This section introduces shortly T-REX architecture but move
%   quickly the focus on a single reactor and gross concepts on the
%   architecture:
%   \begin{itemize}
%   \item For t-trex the whole decision and control problem is reduced
%     around the state variables (or timeline construct). 
%   \item This reduce the execution tracking problem for each reactor on a state
%     identification problem (deduce internal state from the external
%     state), and the control problem to goal posting which will trigger
%     deliberation on the owner of the corresponding timeline. 
%   \item the architecture itself see each reactor as a black box and
%     therefore each reactor can implement its own mechanism in order to
%     resolve both synchronization and deliberation. Still we provide a
%     reactor based on the europa framework that leverage the automated
%     planning capabilities in order to do model based planning and
%     execution with a rich representation of resources and time.
%   \end{itemize}}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
  \caption{\small An example of a \rx reactor hierarchy which is used
    to determine a synchronization order for propagation of values in
    \emph{external} state variables.}
  \label{fig:reactor-DAG}
\end{figure}

As the execution frontier, $\tau$ advances, the \rx agent propagates
the current state information across all reactor state variables
through the {\em synchronization} process.  The sequencing of the
reactors synchronization is identified by traversing a reactor
dependency graph from the least dependent -- ones with no {\em
  external} timelines -- to the most, \ie which typically do not have
{\em internal} timelines. Fig. \ref{fig:reactor-DAG} shows an example
of such a dependency graph implemented as a directed acyclic graph.

@@

For intertwined plan execution, the concept of execution latency is
critical at the agent level since it can impact the
planning/deliberation process on {\em external} state variables of a
reactor. For simplicity we abstract it out as being equivalent to the
reactor latency implying that we assume that all {\em external}
timelines of this reactor have $\lambda_{exec} = 0$. At the \rx agent
level this execution latency is used to define the {\em dispatch
  window} of the same state variable.% : \comment{Needed to say that
  % early on but the flow of this paragraph is not very tight}
%\kcomment{Looks fine to me unless you want to clarify further.}

@@

Requests from a reactor can be {\em recalled} when it is determined
that they are not part of its projected plan anymore. For example if a
reactor needs to replan due to an unexpected event, it will recall all
requests it produced that have not been executed yet.
% \kcomment{replace
%   ``observed'' with ``not incorporated by the owner reactor''?}
% yet
 This recall is sent to the reactor that received the goal(s) to
remove it/them from its current set of objectives.

@@

Both these 
% deliberative \kcomment{``deliberative??''; something else
%   or remove altogether perhaps?} 
functions have different scope and temporal constraints. 
The \texttt{synchronize} call is considered as atomic and will 
focus only on state inference for the current tick by
identifying the current value of {\em internal} state variables from
the latest updates on {\em external} state variables. This process
propagates up in the reactor dependency graph at every tick to ensure
that all reactors are aware of the current state of the
world. \texttt{step} relates to a single decision step of the planning
process focused on the future evolution of state variables; it can
take multiple steps spanning multiple ticks for a reactor to produce a
plan that is both complete and valid. How the reactor identifies that
its plan is complete and valid is its responsibility and depends on
the way each reactor resolves its plan at each {\em step}. The general
idea is that the reactor is able to find a valid strategy in order to
satisfy as many of its {\em internal} goals as possible. For \eu, this
means that the solver is unable to identify any further flaws to be
resolved in the plan.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
stops, its starting point is the last partial plan resulting from
deliberation. This gives an initial frame to the problem that helps to
target the synchronization problem by ensuring that the new {\em
  external} state updates are compatible with this plan. If so,
synchronization propagates these events in the plan and can uses it as
an aid in resolving its {\em current state flaw} by using the plan
information on the exepected {\em internal} state evolution nearby the
execution frontier. 
% \kcomment{``use it
%   as an aid to identify'' what though?}

If they are not compatible, then {\em synchronization} identifies a
plan inconsistency. This results in the retraction of decisions made
during previous {\em deliberation} steps. % in order to
% allow {\em synchronization} to identify its {\em internal} state.
Such a relaxation can be viewed as a backtracking % on the plan search
% triggered exogenously by the world evolution. It is the way our
allowing the reactor to identify that its current plan has failed to
execute triggering further deliberations steps to find an alternate
strategy.

@@

In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
new type of flaw which requires to identify the value of all the {\em
  internal} state variables for the current tick.
% forces identification \kcomment{identification
%   of what? Probably need a little more explanation here.} of internal
% state of the reactor for the current tick. 
This flaw can be defined as:

@@

As the simple example above shows, synchronization involves a
deliberation process with the risk that the search for a solution is
not a straightforward sequence of staright-forward decisions where 
the resolution can be easily identified by the solver as any other
option would leafd immediately to an inonsistency in the plan.
% \kcomment{has a unit decision been explained in Europa?}. 
Since such a process is required by the architecture at \emph{every}
tick for \emph{every} reactor the model needs to be designed carefully.

One critical aspect to take into account is that from a \rx point of
view, synchronization is the only critical task of a reactor. Such a
process is used in order for the agent to identify its current state
and maintain it consistently throughout the agent's lifetime. For a
reactor not being able to find a consistent solution during
synchronization -- implying that after evaluating all possible
solutions provided by a model it was unable to find a consistent plan
-- not only indicates that a reactor is unable to explain the state of
the world, but more importantly it may jeopardize other reactors that
rely on its {\em internal} state. For this reason a failure to
synchronize immediately results in the \rx agent killing a reactor and
notifying any reactor that depends on its {\em internal} state
variables; it does so by posting the special observation
\textsf{Failed}. By so doing we allow for graceful degradation where
other reactors can readapt their plans on receipt of this observation.

@@

When the {\em execution frontier} advances, the reactor planning scope
moves further into the future. This is reflected by the {\em
  deliberation} solver which uses a new deliberation scope one tick
further in the future. As Fig. \ref{fig:plan:window} illustrates, such
a scope allows filtering of tokens that are either necessarily ending
before $\tau$ or necessarily starting after, focusing the planning
problem on only those tokens that are within a ``reasonable'' future.
This filtering window complements the \rx {\em dispatch
  window} from Definition \ref{def:dispatch} which filters out at the
agent level those goals that are before the reactor execution latency
or too far into the future. The planning window reflect this filtering
of the agent by applying to tokens that are generated internally by the
reactor during its deliberation or the {\em goals} for which he
planner decided to postpone after the planning window -- probably
due to other tasks that needed completion beforehand.
% The planning window inside of a reactor
% refines this filtering further, as tokens can be excluded dynamically
% from deliberation as their insertion in the plan and scheduling with
% other tokens generated, can reduce their temporal scope outside of the
% planning scope \kcomment{ this last sentence should be reworked}. 
This allow the planner to focus on its planning horizon and report
decision on flawss thgat are beyond this scope on later ticks. As, 
time advance, these future flaws will eventuallty be part of the
planning window of the reactor and will be then resolved along with
potential new goals received from the agent.
% This helps the deliberation solver reduce the number of tokens to be
% evaluated reducing deliberation cost and increasing performance. As
% time advances, flaws that were initially beyond this temporal horizon
% will become active which a reactor can evaluate.
 This results in an apparent \emph{continuous planning} of the reactor. When there are no
more flaws present in the current planning scope a partial plan
solution is found. This has two effects:

@@

Consider for example the Shopping Agent model where the \texttt{Agent}
is an {\em external} timeline. As the reactor adds a goal to
\texttt{Own} milk and the current {\em external} timeline indicates
that the robot was \texttt{At(home)}, it produces a plan as shown in
Fig. \ref{fig:shop:exec0}. The {\em external} part of the plan --
namely the tokens \texttt{Going(home, superMarket)} and
\texttt{At(SuperMarket)} -- are installed as goals for the reactor
managing this state variable internally.

\begin{figure}[!htb]
  \centering
  \subfloat[\small A plan produced before synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.45\columnwidth]{figs/shoping_exec_t0}}
  \subfloat[\small Result of synchronization after the {\em Going}
  observation was received by the reactor with the plan showed in
    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=0.48\columnwidth]{figs/shoping_exec_t1}}
    \caption{\small Shopping example: Tokens in red indicate
      observations, tokens in blue the goal that this plan attempts to
      solve for. Arrows indicate temporal constraints. }
  \label{fig:shop:exec}
\end{figure}

>>>>

\section{Planning and execution with \rx}
\label{sec:arch}

@@

% \begin{figure}[!htbp]
%   \centering
%   \includegraphics[scale=0.35]{figs/DAG-hierarchy}
%   \caption{\small An example of a \rx reactor hierarchy which is used
%     to determine a synchronization order for propagation of values in
%     \emph{external} state variables.}
%   \label{fig:reactor-DAG}
% \end{figure}
% \fcomment{I removed this figure as the next one is mucjh more
%   detailed and vehiculate the same idea}

As the execution frontier, $\tau$ advances, the \rx agent propagates
the current state information across all reactor state variables
through the {\em synchronization} process.  The sequencing of the
reactors synchronization is identified by traversing a reactor
dependency graph from the least dependent -- ones with no {\em
  external} state variables -- to the most, \ie which typically do not have
{\em internal} state variables. Fig. \ref{fig:agent} shows an example
of such a dependency graph implemented as a directed acyclic graph.

@@

For intertwined plan execution, the concept of execution latency is
critical at the agent level since it can impact the
planning/deliberation process on {\em external} state variables of a
reactor. For simplicity we abstract it out as being equivalent to the
reactor latency implying that we assume that all {\em external}
timelines of this reactor have $\lambda_{exec} = 0$. At the \rx agent
level this execution latency is used to define the {\em dispatch
  window} of the same state variable.

@@

Requests from a reactor can be {\em recalled} when it is determined
that they are not part of its projected plan anymore. For example if a
reactor needs to replan due to an unexpected event, it will recall all
requests produced that have yet to be executed.  This recall is also
sent to the reactor(s) that received the goal(s) to remove it/them
from its current set of objectives.

@@

Both these functions have different scope and temporal constraints.
The \texttt{synchronize} call is considered as atomic and will focus
only on state inference for the current tick by identifying the
current value of {\em internal} state variables from the latest
updates on {\em external} state variables. This process propagates up
in the reactor dependency graph at every tick to ensure that all
reactors are aware of the current state of the world. \texttt{step}
relates to a single decision step of the planning process focused on
the future evolution of state variables; it can take multiple steps
spanning multiple ticks for a reactor to produce a plan that is both
complete and valid. How the reactor identifies that its plan is
complete and valid is its responsibility and depends on the way each
reactor resolves its plan at each {\em step}. The general idea is that
the reactor is able to find a valid strategy in order to satisfy as
many of its {\em internal} goals as possible. For \eu, this means that
the solver is unable to identify any further flaws to be resolved in
the plan.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
stops, its starting point is the last partial plan resulting from
deliberation. This gives an initial frame to the problem that helps to
target the synchronization problem by ensuring that the new {\em
  external} state updates are compatible with this plan. If so,
synchronization propagates these events in the plan and can use it as
an aid in resolving its {\em current state flaw}. It does so by using
the plan information on the exepected {\em internal} state evolution
near the execution frontier.

If they are not compatible, then {\em synchronization} identifies a
plan inconsistency. This results in the retraction of decisions made
during previous {\em deliberation} steps.  Such a relaxation can be
viewed as a backtracking allowing the reactor to identify that its
current plan has failed to execute triggering further deliberations
steps to find an alternate strategy.

@@

In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
new type of flaw which requires identification of all values of {\em
  internal} state variables for the current tick. This flaw can be
defined as:

@@

As the example above shows, synchronization involves a deliberation
process. The risk is such that the search for a solution is not a
simple sequence of decisions where the resolution can be easily
identified by the solver as any other option would lead immediately to
n plan inconsistency. Since such a process is required by the \rx at
\emph{every} tick for \emph{every} reactor the model needs to be
designed carefully.

One critical aspect to take into account is that from a \rx point of
view, synchronization is the only critical task of a reactor. Such a
process is used in order for the agent to identify its current state
and maintain it consistently throughout the agent's lifetime. For a
reactor not being able to find a consistent solution during
synchronization -- implying that after evaluating all possible
solutions provided by a model it was unable to find a consistent plan
-- not only indicates that a reactor is unable to explain the state of
the world, but more importantly it may jeopardize other reactors that
rely on its {\em internal} state. For this reason a failure to
synchronize immediately results in the \rx agent killing a reactor and
notifying any reactor that depends on its {\em internal} state
variables; it does so by posting the special observation
\textsf{Failed}. This special state is a special predicate which is
acceptable for all the {\em internal} and {\em external} timelines of
a \eu reactor.  By so doing we allow for graceful degradation where
other reactors can readapt their plans on receipt of this observation.

@@

When the {\em execution frontier} advances, the reactor planning scope
moves further into the future. This is reflected by the {\em
  deliberation} solver which uses a new deliberation scope one tick
further in the future. As Fig. \ref{fig:plan:window} illustrates, such
a scope allows filtering of tokens that are either necessarily ending
before $\tau$ or necessarily starting after, focusing the planning
problem on only those tokens that are within a ``reasonable'' future.
This filtering window complements the \rx {\em dispatch window} from
Definition \ref{def:dispatch} which filters out at the agent level
those goals that are before the reactor execution latency or too far
into the future.  The planning window refines this filtering further,
as tokens can be excluded dynamically from deliberation as their
insertion in the plan and scheduling with other generated tokens can
alter their temporal scope to be outside of the planning window.  This
allows the planner to focus on its planning horizon and report
decision on flaws that are beyond this scope. As, time advances, these
potential future flaws will eventually be part of the planning window
of the reactor and will be then resolved along with potentially new
goals.  This in turn, results in an apparent \emph{continuous
  planning} of the reactor. When there are no more flaws present in
the current planning scope a partial plan solution is found. This has
two effects:

@@

Consider for example the Shopping Agent model where the timeline 
\texttt{Agent.position} is {\em external} while \texttt{Agent} is
implemented a an {\em internal} timeline. As the reactor adds a goal to
\texttt{Own} milk and the current {\em external} timeline indicates
that the robot was \texttt{At(home)}, it produces a plan as shown in
Fig. \ref{fig:shop:exec0}. The {\em external} part of the plan --
namely the tokens \texttt{Going(home, superMarket)} and
\texttt{At(SuperMarket)} -- are installed as goals for the reactor
managing this state variable internally.
x1
\begin{figure}[!htb]
  \centering
  \subfloat[\small A plan produced before
  synchronization.]{\label{fig:shop:exec0}\includegraphics[width=0.7\columnwidth]{figs/shoping_exec_t0}} \\
  \subfloat[\small Result of synchronization after the {\em Going}
  observation was received by the reactor with the plan showed in
    \ref{fig:shop:exec0}.]{\label{fig:shop:exec1}\includegraphics[width=.7\columnwidth]{figs/shoping_exec_t1}}
    \caption{\small Shopping example: Tokens in red indicate
      observations, tokens in blue the goal that this plan attempts to
      solve for. Arrows indicate temporal constraints. }
  \label{fig:shop:exec}
\end{figure}

### 2012-02-23 trunk/related.tex
Rewrote some parts of related.tex with comments

There are costs associated with such inference based control are also
not inconsequential. Foremost among them is the steep learning curve
that comes with the design of the domain model. With limited support
tools necessary for knowledge capture and design of the constraints
often in stylized higher-level languages (see \cite{NDDL} for an
example), it requires the modeler to think in alternative ways to do
problem solving. More importantly it is often necessary for the
modeler to be exposed to the internal representation of the planner
and how it performs search. THe level of expertise is often well above
what would be expected of a typical well-rounded Computer Science
graduate student.

However this is mild criticism given the end benefits and the
scientific and engineering goals associated with adaptive and
persistent observation for marine robots whether for civil or military
applications. Given the general problem-solving nature of such
deliberative systems, there is also a substantial overall benefit for
the field of robotics.

@@

Planning and plan execution are not new to the overall field of
robotics. Early motivation of building computational mechanisms for
decision making were intended to be deployed on robotic devices by
very definition. The seminal volume \cite{computersthought} very
clearly articulates how physical manifestation of robots could be
controlled for task planning and execution. The very first planner
\cite{green69} was an exercise in computational theorem proving with
early state space planners popularized by \texttt{STRIPS}
\cite{strips71} dominating the academic landscape. An early
demonstration of planning as a situated agent was the highly
influential \texttt{Shakey} the robot experiment
\cite{shakey84}. Subsequent work driven largely by defense funding in
the United States sequestered the discipline in academic labs
principally applied for algorithmic development and if embedded then
in testing within the confines of modest indoor settings. The leap to
applying them for real world problem solving is relatively new and
driven largely by NASA applications pioneered by \cite{mus94,mus98,
  jonsson00, rajan00, chien05, bresina05}.

The dominant approach for building agent control systems utilize a
$3$-layered architecture (see \cite{gat98} for a well thought out
rationale), notable examples of which include \texttt{IPEM}
\cite{AmbrosIngerson88}, \texttt{ROGUE} \cite{Haigh98}, the LAAS
Architecture \cite{alami:1998p820}, the Remote Agent Experiment
\texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). Early
efforts were cognizant of if absent with, dealing with time and
resources for such experiments.

@@

\subsection{The Remote Agent Experiment and beyond}

@@

Our system, \texttt{T-REX}, the focus of this chapter, is an agent
control structure formally defined as a composition of coordinated
control loops with a recurring sense-plan-act (SPA) cycle with
deliberation at its core. \texttt{T-REX} was designed after carefully
evaluating lessons learned from \texttt{IDEA} and owes substantially
to the notion of partitioned problem solving from that effort. However
this partitioning is systematic and methodologically principled where
we manage the information flow within the partitioned structure to
ensure consistency in order to direct the flow of goals and
observations in a timely manner. The resulting control structure
improves scalability since many details of each controller can be
encapsulated within a single control loop.  Furthermore, partitioning
increases robustness since controller failure can be localized to
enable graceful system degradation, making this an effective
\emph{divide-and-conquer} approach to the overall control problem.

>>>>

There are costs associated with such inference based control which are 
not inconsequential. Foremost among them is the steep learning curve
that comes with the design of the domain model. With limited support
tools necessary for knowledge capture and design of the constraints
often in stylized higher-level languages (see \cite{NDDL} for an
example), it requires the modeler to think in alternative ways to do
problem solving. More importantly it is often necessary for the
modeler to be exposed to the internal representation of the planner
and how it performs search. The level of expertise is often well above
what would be expected of a typical well-rounded Computer Science
graduate student.

However this is mild criticism given the end benefits and the
scientific and engineering goals associated with adaptive and
persistent observation for marine robots whether for civil or military
applications. Given the general problem-solving nature of such
deliberative systems, there is also a substantial overall benefit for
the field of robotics. 
\fcomment{Though it woulfd  be usefull to emphasize this ...}
The cost of modeling put aduring the intial design, for example,
should be balanced against the cost of redesigning
a new mission script eate every new deployment in more ``classical''
script-based controllers.

@@

Planning and plan execution are not new to the overall field of
robotics. Early motivation of building computational mechanisms for
decision making were intended to be deployed on robotic devices by
very definition. The seminal volume \cite{computersthought} very
clearly articulates how physical manifestation of robots could be
controlled by task planning and execution. The very first planner
\cite{green69} was an exercise in computational theorem proving with
early state space planners popularized by \texttt{STRIPS}
\cite{strips71} dominating the academic landscape. An early
demonstration of planning as a situated agent was the highly
influential \texttt{Shakey} the robot experiment
\cite{shakey84}. Subsequent work driven largely by defense funding in
the United States sequestered the discipline in academic labs
principally applied for algorithmic development and if embedded then
in testing within the confines of modest indoor settings. The leap to
applying them for real world problem solving is relatively new and
driven largely by NASA applications pioneered by \cite{mus94,mus98,
  jonsson00, rajan00, chien05, bresina05}.

The dominant approach for building agent control systems utilize a
$3$-layered architecture (see \cite{gat98} for a well thought out
rationale), notable examples of which include \texttt{IPEM}
\cite{AmbrosIngerson88}, \texttt{ROGUE} \cite{Haigh98}, the LAAS
Architecture \cite{alami:1998p820}, the Remote Agent Experiment
\texttt{RAX} \cite{mus98} and the Autonomous Spacecraft Experiment
\texttt{ASE} \cite{chien99} (see \cite{Knight01} for a survey). Early
efforts were cognizant of if absent with, dealing with time and
resources for such experiments.\fcomment{This sentence does not make
  sence to me}

@@

\subsection{The Remote Agent Experiment and beyond}
\label{sec:rabeyond}

@@

Our system, \texttt{T-REX}, the focus of this chapter, is an agent
control structure formally defined as a composition of coordinated
control loops where sensing, planning and acting are resulting from 
concurrent deliberating tasks within a formal framework. \fcomment{Got
  rid of sense-plan-act reference here. \rx plan/sense while sensing/planning.}
\texttt{T-REX} was designed after carefully
evaluating lessons learned from \texttt{IDEA} and owes substantially
to the notion of partitioned problem solving from that effort. However
this partitioning is systematic and methodologically principled where
we manage the information flow within the partitioned structure to
ensure consistency in order to direct the flow of goals and
observations in a timely manner. The resulting control structure
improves scalability since many details of each controller can be
encapsulated within a single control loop.  Furthermore, partitioning
increases robustness since controller failure can be localized to
enable graceful system degradation, making this an effective
\emph{divide-and-conquer} approach to the overall control problem.
